---
title: "DSA 557 1.0 Scientific Writing and Communication"
author: 
  - "Dr Thiyanga S. Talagala"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#253494",
  secondary_color = "#c7e9b4",
  inverse_header_color = "#a1d99b"
)
```

<style type="text/css">
.remark-slide-content {
    font-size: 30px;
    padding: 1em 4em 1em 4em;
}
</style>

# What is Scientific Writing and Communication?


Scientific writing and communication refer to the skills and practices involved in effectively conveying scientific information to various audiences. 

---

# Importance of scientific writing

1. Sharing your research

2. Reproducibility

3. Funding and career advancement

3. Public engagement

---

# Types of Scientific Documents

## Where do we find scientific writing?

---

# Practical tips and tools?

- Use simple, straight forward language

- Use clear headings and subheadings

- Always revise your work and seek feedback

- Use a plagiarism checker


---
## Papers selected for discussion today

1. [A Layered Grammar of Graphics](https://www.tandfonline.com/doi/epdf/10.1198/jcgs.2009.07098?needAccess=true)

2. [Forecasting: theory and practice](https://www.sciencedirect.com/science/article/pii/S0169207021001758)

3. [Forecasting with temporal hierarchies](https://www.sciencedirect.com/science/article/abs/pii/S0377221717301911)

4. [Meta-learning how to forecast time series](https://onlinelibrary.wiley.com/doi/abs/10.1002/for.2963)

5. [A Hexagon Tile Map Algorithm for Displaying Spatial Data](https://journal.r-project.org/articles/RJ-2023-021/)


6. [Calendar-based graphics for visualizing...](https://pdf.earo.me/calendar-vis.pdf)

7. [A Projection Pursuit Forest Algorithm for Supervised Classification](https://www.tandfonline.com/doi/epdf/10.1080/10618600.2020.1870480?needAccess=true)

---

# Research papers/ Journal Article /Manuscript (before publication)/ Scholarly Article

---

# Review the articles shared today
# What writing structures do they follow?

---

# Examine the articles shared today

# In what ways do their purposes and contents differ?

---

# Selecting good papers

In-class demo

---

## Writing introduction

**Hook**

A compelling opening sentence or idea to grab the reader's attention.

---
## Types of hooks

**1. Statistic or Fact Hook**

Start with a surprising or impactful statistic or fact related to your topic.

Example:

Over 80% of the world’s wastewater is released into the environment untreated, affecting millions of people daily.

---

## Types of hooks (cont.)

**2. Question Hook**

Begin with a thought-provoking question to engage the reader.

---
## Types of hooks (cont.)

**3. Quotation Hook**

Use a relevant and powerful quote from a well-known figure or authoritative source.

Example:

In 1976, a British statistician named George Box wrote the famous line, “All models are wrong, some are useful.

---
## Types of hooks (cont.)

**4. Anecdotal Hook**

Start with a brief personal story or anecdote that illustrates the issue.

Example:

When I first visited the rural village, I noticed children walking miles to access clean water—a reality for millions across the globe.

---
## Types of hooks (cont.)

**5. Historical Hook**

Begin by referencing a historical event or background related to the topic.

Example:

In 1854, John Snow mapped a cholera outbreak in London—marking the birth of modern epidemiology. Today, we face similar challenges with disease mapping in developing nations.

---
## Types of hooks (cont.)

**6. Contrasting Statement Hook**

Present a contradiction or surprising contrast to pique curiosity.

Example:

Despite technological advancements, rural students in many countries still lack access to basic educational resources.

---
## Types of hooks (cont.)

**7. Definition Hook**

Define a term or concept that is central to your topic, especially if it has varied meanings.

Example:

Machine learning interpretability refers to the degree to which a human can understand the reasoning behind a machine learning model's predictions.

---
## Other components - Introduction


1. Background / Context

2. Problem Statement

3. Review of Key Literature (optional in short papers)

4. Research Aim / Objective

5. Research Questions or Hypotheses (if applicable)

6. Significance of the Study

7. Overview of the Paper / Structure 

---

## Types of Background in a Research Introduction

1. Thematic Background

2. Contextual Background

3. Conceptual Background

4. Methodological Background

5. Problem-Oriented Background

6. Literature-Based Background

7. Policy or Practice-Based Background




---
## Scientific writing with overleaf

Sign up at https://www.overleaf.com/

---

# Lecture 2
---
#  Different structures to break down the background section

---

## Application work

- Introduction to the Practical Problem


- Existing Methods / Technologies



- Gap in Application / Practice



- What This Paper Does (Your Approach)



- Significance of the Study



- Structure of the Paper

---

## Theoretical Papers

- Introduction to the Theoretical Area


- Review of Key Existing Theories/Models


- Theoretical Gap / Unresolved Issues

- What You Do (Main Theoretical Contribution)


- Significance of the Contribution


- Structure of the Paper

---

## Review Papers


- Context and Motivation for the Review

- Scope of the Review

- Existing Reviews and Their Limitations

- What You Do in This Paper

- Significance of the Review


- Structure of the Paper

---

# Topic sentence 

Topic sentence is the first sentence of a paragraph that introduces the main idea or focus of that paragraph

----

# Topic sentence: Purpose

- It tells the reader what the paragraph is about.

- It acts as a guide for both the writer and the reader.

- It helps with clarity, flow, and coherence in your writing.

---
## A good topic sentence is:

- Clear and specific (not too broad or vague)

- Often connects to the previous paragraph (transition)

- Often introduces a claim or idea that the rest of the paragraph explains or supports

---
## Hook vs. Topic Sentence

In-class discussion

---

## Transition strategies

---



- Introduction → Gap in Knowledge

    "However, despite these advances, several challenges remain."

- Gap → What You Do

    "To address this gap, we propose..."

- What You Do → Significance

    "This contribution is significant because..."

- Significance → Structure of the Paper

    "The remainder of this paper is organized as follows:"